A Data-Driven Framework for Neural Field Modeling

### 2010-12-29 ltx/first_order_framework/LaTeX_Manuscript.tex
removed some mistakes from correation analyses

The mapping between the membrane voltage and the electrophysiological data, denoted by $\mathbf y_t$, is modelled using the observation function that incorporates sensors with a spatial extent by
\begin{equation}
    \label{eq:ObservationEquation}
	y_t(\mathbf{r}_n) =
	\int_{\Omega}{
	    m\left(\mathbf{r}_n-\mathbf{r}'\right)v_t\left(\mathbf{r}'\right)
	\, d\mathbf{r}'} + 
	\varepsilon_t(\mathbf{r}_n), 
\end{equation}
\dean{(dean: I added $(\mathbf{r}_n)$ to index the observations)}
where $m\left(\mathbf{r}_n-\mathbf{r}'\right)$ is the observation kernel, $\mathbf{r}_n$ defines the location of the electrodes in the field, $n=1,...,n_y$ indexes the sensors and $\varepsilon_t(\mathbf{r}_n) \sim \mathcal{N}\left(0,\boldsymbol{\Sigma}_{\varepsilon}\right)$ denotes a multivariate normal distribution with mean zero and covariance matrix $\boldsymbol{\Sigma}_{\varepsilon}$.

@@

The first term on the right hand side of equation~\ref{eq:spatialxcorr} is
\begin{equation}
	\xi\mathbf{E}\left[\bar{y}_t\left(\mathbf{r}_n + \boldsymbol{\tau}\right) \int_{\Omega}{ m\left(\mathbf{r}_n-\mathbf{r}'\right) v_t\left(\mathbf{r}\right) d\mathbf{r}'} \right] = \xi R_{\bar{y}_{t},\bar{y}_{t}}(\boldsymbol{\tau}),
\end{equation}
which is the spatial autocorrelation of the observations scaled by the decay parameter $\xi$. DOES THIS GIVE US AN ESTIMATE OF THE FIELD DISTURBANCE COVARIANCE? NUMERICALLY IT SEEMS TO GIVE SOMETHING VERY CLOSE. DEPENDS ON THE BANDWIDTH OF M. The last term on the right hand side of equation~\ref{eq:spatialxcorr} equates to zero since the disturbance added to the field at time $t$ and assumed to be uncorrelated to the field at time $t$.

To understand the middle term, the sigmoid activation function, $f(\cdot)$, is approximated by the piecewise function
\begin{equation}
	\hat{f}(v_t(\mathbf{r}')) = \left\{ \begin{array}{ll}
		0, & v_t(\mathbf{r}') \le v_1 \\
		\varsigma v_t(\mathbf{r}'), &  v_1 < v_t(\mathbf{r}') < v_2 \\
		1, & v_t(\mathbf{r}') \ge v_2 \\ 
		\end{array}\right.
\end{equation}

Evaluating the middle term for the case where the field is in the linear region of the nonlinearity gives
\begin{align}
	middle term &= T_s\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ m\left(\mathbf{r}_n-\mathbf{r}'\right)\int_\Omega { w\left(\mathbf{r}',\mathbf{r}''\right) f\left(v_t\left(\mathbf{r}''\right)\right)\, d\mathbf{r}''}\, d\mathbf{r}'}\right]  \\
	&=T_s\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ m\left(\mathbf{r}_n-\mathbf{r}'\right)\int_\Omega { w\left(\mathbf{r}',\mathbf{r}''\right) \varsigma v_t\left(\mathbf{r}''\right)\, d\mathbf{r}''}\, d\mathbf{r}'}\right] \label{eq:preswap}  \\
	&= T_s\varsigma\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ w\left(\mathbf{r}_n,\mathbf{r}'\right) \int_\Omega {  m\left(\mathbf{r}'-\mathbf{r}''\right) v_t\left(\mathbf{r}''\right)\, d\mathbf{r}''}\, d\mathbf{r}'}\right]  \label{eq:postswap} \\
	&= T_s\varsigma\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ w\left(\mathbf{r}_n,\mathbf{r}'\right) \bar{y}_t\left( \mathbf{r}' \right)\, d\mathbf{r}'}\right],	
\end{align}
where the associativity and commutativity algebraic properties of the spatial convolution where used to move from equation~\ref{eq:preswap} to equation~\ref{eq:postswap}.

Evaluating the middle term for the saturation regions. If $v_t(\mathbf{r}') \le v_1$ then the middle term will equate to zero. If $v_t(\mathbf{r}') \ge v_2$ then we get
\begin{align}
	middle term &= T_s\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ m\left(\mathbf{r}_n-\mathbf{r}'\right)\int_\Omega { w\left(\mathbf{r},\mathbf{r}''\right)\, d\mathbf{r}''}\, d\mathbf{r}'}\right] \\
	&= T_s a_1 \mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ m\left(\mathbf{r}_n - \mathbf{r}'\right)\, d\mathbf{r}'}\right] \\
	&= T_s a_1 a_2 \mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \right],
\end{align}
where $a_1$ is the integral of the connectivity kernel and $a_2$ is the integral of the observation kernel.

Under the assumption that the field is in the linear region of the sigmoid for the majority of time then
\begin{equation}
	R_{y_{t},y_{t+1}}(\boldsymbol{\tau}) \approx \xi R_{  \bar{y}_{t},\bar{y}_{t} } (\boldsymbol{\tau}) + T_s\varsigma\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ w\left(\mathbf{r}_n,\mathbf{r}'\right) \bar{y}_t\left( \mathbf{r}' \right)\, d\mathbf{r}'}\right].
\end{equation}
The autocorrelation of the noisy observations is
\begin{equation}
	R_{y_{t},y_{t}}(\boldsymbol{\tau}) = R_{\bar{y}_{t},\bar{y}_{t} } (\boldsymbol{\tau}) + \delta(\boldsymbol{\tau})\sigma_{\varepsilon}^2
\end{equation}

% Therefore we can write
% \begin{equation}
% 	R_{v_{t},v_{t+1}}(\mathbf{\tau}) = \xi R_{v_{t},v_{t}}(\mathbf{\tau}) + T_s R_{v_{t},g_{t}}(\mathbf{\tau})
% \end{equation}
% and
% \begin{equation}\label{eq:xcorr_firing_and_field}
% 	T_s R_{v_{t},g_{t}}(\mathbf{\tau}) = R_{v_{t},v_{t+1}}(\mathbf{\tau})-\xi R_{v_{t},v_{t}}(\mathbf{\tau}).
% \end{equation}
% The spatial cross-correlation function $R_{v_{t},g_{t}}(\mathbf{\tau})$ depends on the shape of the spatial mixing kernel $w\left( \mathbf{r},\mathbf{r}' \right)$. Furthermore, the sign (polarity) of $R_{v_{t},g_{t}}(\mathbf{\tau})$ will be the same as the $w\left( \mathbf{r},\mathbf{r}' \right)$ for a given spatial location since the sigmoid function $f\left( v_t\left( \mathbf{r}' \right) \right) > \mathbf{0}$. Therefore, the support of the kernel can be inferred given the cross-correlation function $R_{v_{t},g_{t}}(\mathbf{\tau})$. The right hand side of equation~\ref{eq:xcorr_firing_and_field} can be calculated from data if $\xi$ is known. However, since $\xi$ is a decay parameter, we know that $\xi \in (0,1)$. Also, if we assume that
% \begin{equation}
% 	\xi R_{v_{t},v_{t}}(\mathbf{\tau}) \gg T_s R_{v_{t},g_{t}}(\mathbf{\tau})
% \end{equation}
% and
% \begin{equation}
% 	R_{v_{t},g_{t}}(\mathbf{0}) > 0,
% \end{equation}
% which corresponds to local excitation, then we can guess $\xi$ by incrementally increasing it from zero and evaluating equation~\ref{eq:xcorr_firing_and_field} until $T_s R_{v_{t},g_{t}}(\mathbf{0}) < 0$. The value of $\xi$ before $T_s R_{v_{t},g_{t}}(\mathbf{0}) < 0$ will provide the approximation.

>>>>

The mapping between the membrane voltage and the electrophysiological data, denoted by $\mathbf y_t$, is modelled using the observation function that incorporates sensors with a spatial extent by
\begin{equation}
    \label{eq:ObservationEquation}
	y_t(\mathbf{r}_n) =
	\int_{\Omega}{
	    m\left(\mathbf{r}_n-\mathbf{r}'\right)v_t\left(\mathbf{r}'\right)
	\, d\mathbf{r}'} + 
	\varepsilon_t(\mathbf{r}_n), 
\end{equation}
where $m\left(\mathbf{r}_n-\mathbf{r}'\right)$ is the observation kernel, $\mathbf{r}_n$ defines the location of the electrodes in the field, $n=1,...,n_y$ indexes the sensors and $\varepsilon_t(\mathbf{r}_n) \sim \mathcal{N}\left(0,\boldsymbol{\Sigma}_{\varepsilon}\right)$ denotes a multivariate normal distribution with mean zero and the covariance matrix $\boldsymbol{\Sigma}_{\varepsilon} = \sigma_{\varepsilon}^2\mathbf{I}$.

@@

The first term on the right hand side of equation~\ref{eq:spatialxcorr} is
\begin{equation}
	\xi\mathbf{E}\left[\bar{y}_t\left(\mathbf{r}_n + \boldsymbol{\tau}\right) \int_{\Omega}{ m\left(\mathbf{r}_n-\mathbf{r}'\right) v_t\left(\mathbf{r}\right) d\mathbf{r}'} \right] = \xi R_{\bar{y}_{t},\bar{y}_{t}}(\boldsymbol{\tau}),
\end{equation}
which is the spatial autocorrelation of the observations scaled by the decay parameter $\xi$. The last term on the right hand side of equation~\ref{eq:spatialxcorr} equates to zero since the disturbance added to the field at time $t$ and assumed to be uncorrelated to the field at time $t$.

To understand the middle term, the sigmoid activation function, $f(\cdot)$, is approximated by the piecewise function
\begin{equation}
	\hat{f}(v_t(\mathbf{r}')) = \left\{ \begin{array}{ll}
		0, & v_t(\mathbf{r}') \le v_1 \\
		\varsigma v_t(\mathbf{r}'), &  v_1 < v_t(\mathbf{r}') < v_2 \\
		1, & v_t(\mathbf{r}') \ge v_2 \\ 
		\end{array}\right.
\end{equation}
Evaluating the middle term of the RHS of equation~\ref{eq:spatialxcorr} for the case where the field is in the linear region of the nonlinearity gives
\begin{align}
	middle term &= T_s\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ m\left(\mathbf{r}_n-\mathbf{r}'\right)\int_\Omega { w\left(\mathbf{r}',\mathbf{r}''\right) f\left(v_t\left(\mathbf{r}''\right)\right)\, d\mathbf{r}''}\, d\mathbf{r}'}\right]  \\
	&=T_s\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ m\left(\mathbf{r}_n-\mathbf{r}'\right)\int_\Omega { w\left(\mathbf{r}',\mathbf{r}''\right) \varsigma v_t\left(\mathbf{r}''\right)\, d\mathbf{r}''}\, d\mathbf{r}'}\right] \label{eq:preswap}  \\
	&= T_s\varsigma\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ w\left(\mathbf{r}_n,\mathbf{r}'\right) \int_\Omega {  m\left(\mathbf{r}'-\mathbf{r}''\right) v_t\left(\mathbf{r}''\right)\, d\mathbf{r}''}\, d\mathbf{r}'}\right]  \label{eq:postswap} \\
	&= T_s\varsigma\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ w\left(\mathbf{r}_n,\mathbf{r}'\right) \bar{y}_t\left(\mathbf{r}'\right)\, d\mathbf{r}'}\right],	
\end{align}
where the associativity and commutativity algebraic properties of the spatial convolution where used to move from equation~\ref{eq:preswap} to equation~\ref{eq:postswap}.
% 
% Evaluating the middle term for the saturation regions. If $v_t(\mathbf{r}') \le v_1$ then the middle term will equate to zero. If $v_t(\mathbf{r}') \ge v_2$ then we get
% \begin{align}
% 	middle term &= T_s \mathbf{E} \left[ \bar{y}_t\left(\mathbf{r}_n+\boldsymbol{\tau}\right) \int_{\Omega} { m\left(\mathbf{r}_n-\mathbf{r}'\right) \int_\Omega { w\left(\mathbf{r},\mathbf{r}''\right) \, d\mathbf{r}''} \, d\mathbf{r}'} \right] \\
% 	&= T_s a_1 \mathbf{E}\left[ \bar{y}_t \left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ m\left(\mathbf{r}_n - \mathbf{r}'\right)\, d\mathbf{r}'} \right] \\
% 	&= T_s a_1 a_2 \mathbf{E}\left[ \bar{y}_t \left( \mathbf{r}_n + \boldsymbol{\tau} \right) \right],
% \end{align}
% where $a_1$ is the integral of the connectivity kernel and $a_2$ is the integral of the observation kernel.

Under the assumption that the field is in the linear region of the sigmoid for the majority of time then
\begin{equation}\label{eq:approx_xcorr}
	R_{y_{t},y_{t+1}}(\boldsymbol{\tau}) \approx \xi R_{  \bar{y}_{t},\bar{y}_{t} } (\boldsymbol{\tau}) + T_s\varsigma\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ w\left(\mathbf{r}_n,\mathbf{r}'\right) \bar{y}_t\left( \mathbf{r}' \right)\, d\mathbf{r}'}\right].
\end{equation}
The autocorrelation of the noisy observations is
\begin{equation}\label{eq:autocorrelation_of_y}
	R_{y_{t},y_{t}}(\boldsymbol{\tau}) = R_{\bar{y}_{t},\bar{y}_{t} } (\boldsymbol{\tau}) + \delta(\boldsymbol{\tau})\sigma_{\varepsilon}^2,
\end{equation}
where $\delta(\cdot)$ is the Kronecker delta function. Rearranging equation~\ref{eq:autocorrelation_of_y} and substituting it into equation~\ref{eq:approx_xcorr} gives
\begin{equation}
	R_{y_{t},y_{t+1}}(\boldsymbol{\tau}) \approx \xi (R_{y_{t},y_{t}}(\boldsymbol{\tau}) - \delta(\boldsymbol{\tau}) \sigma_{\varepsilon}^2) + T_s\varsigma\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ w\left(\mathbf{r}_n,\mathbf{r}'\right) \bar{y}_t\left( \mathbf{r}' \right)\, d\mathbf{r}'}\right],	
\end{equation}
and
\begin{equation}\label{eq:support_estimator}
	R_{y_{t},y_{t+1}}(\boldsymbol{\tau}) - \xi (R_{y_{t},y_{t}}(\boldsymbol{\tau}) - \delta(\boldsymbol{\tau}) \sigma_{\varepsilon}^2) \approx T_s\varsigma\mathbf{E}\left[ \bar{y}_t\left( \mathbf{r}_n + \boldsymbol{\tau} \right) \int_{\Omega}{ w\left(\mathbf{r}_n,\mathbf{r}'\right) \bar{y}_t\left( \mathbf{r}' \right)\, d\mathbf{r}'}\right].	
\end{equation}
The right hand side of equation~\ref{eq:support_estimator} will have the structure of the connectivity kernel.

This may hold:
\begin{equation}
	R_{\bar{y}_{t},\bar{y}_{t} } (\boldsymbol{\tau}) = \sigma_d \int_{\Omega}{ m(\mathbf{r}_n+\boldsymbol{\tau}-\mathbf{r}')\int_{\Omega}{ m(\mathbf{r}'_n-\mathbf{r}'') \gamma(\mathbf{r}'-\mathbf{r}'') d\mathbf{r}''} d\mathbf{r}'}.
\end{equation}
If it does then we can estimate the support of the disturbance using the correlation analysis if the bandwidth of the sensor is wider than the bandwidth of the disturbance.
% Therefore we can write
% \begin{equation}
% 	R_{v_{t},v_{t+1}}(\mathbf{\tau}) = \xi R_{v_{t},v_{t}}(\mathbf{\tau}) + T_s R_{v_{t},g_{t}}(\mathbf{\tau})
% \end{equation}
% and
% \begin{equation}\label{eq:xcorr_firing_and_field}
% 	T_s R_{v_{t},g_{t}}(\mathbf{\tau}) = R_{v_{t},v_{t+1}}(\mathbf{\tau})-\xi R_{v_{t},v_{t}}(\mathbf{\tau}).
% \end{equation}
% The spatial cross-correlation function $R_{v_{t},g_{t}}(\mathbf{\tau})$ depends on the shape of the spatial mixing kernel $w\left( \mathbf{r},\mathbf{r}' \right)$. Furthermore, the sign (polarity) of $R_{v_{t},g_{t}}(\mathbf{\tau})$ will be the same as the $w\left( \mathbf{r},\mathbf{r}' \right)$ for a given spatial location since the sigmoid function $f\left( v_t\left( \mathbf{r}' \right) \right) > \mathbf{0}$. Therefore, the support of the kernel can be inferred given the cross-correlation function $R_{v_{t},g_{t}}(\mathbf{\tau})$. The right hand side of equation~\ref{eq:xcorr_firing_and_field} can be calculated from data if $\xi$ is known. However, since $\xi$ is a decay parameter, we know that $\xi \in (0,1)$. Also, if we assume that
% \begin{equation}
% 	\xi R_{v_{t},v_{t}}(\mathbf{\tau}) \gg T_s R_{v_{t},g_{t}}(\mathbf{\tau})
% \end{equation}
% and
% \begin{equation}
% 	R_{v_{t},g_{t}}(\mathbf{0}) > 0,
% \end{equation}
% which corresponds to local excitation, then we can guess $\xi$ by incrementally increasing it from zero and evaluating equation~\ref{eq:xcorr_firing_and_field} until $T_s R_{v_{t},g_{t}}(\mathbf{0}) < 0$. The value of $\xi$ before $T_s R_{v_{t},g_{t}}(\mathbf{0}) < 0$ will provide the approximation.
